\PIII, linearizing unit cell parameters

### 2025-08-27 P3/P3.tex
CommandUsage --  add MATRIXORDER to the help. P3, 3D_Niggli_Matrices, and LatticeMatching_2025 are paper in progress.

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}

@@

@@

% Crystallographic space macros
\newcommand{\HVI}{\ensuremath{\mathbf{H}^{6}}}
\newcommand{\GVI}{\ensuremath{\mathbf{G}^{6}}}
\newcommand{\SVI}{\ensuremath{\mathbf{S}^{6}}}
\newcommand{\PIII}{\ensuremath{\mathbf{P}^{3}}}
\newcommand{\FIII}{\ensuremath{\mathbf{F}^{3}}}
\newcommand{\CIII}{\ensuremath{\mathbf{C}^{3}}}
\newcommand{\RIII}{\ensuremath{\mathbb{R}^{3}}}

@@

	\title{\PIII, a simple metric for \HVI}
	\shorttitle{properties of \PIII}

@@

	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}

@@

	\begin{abstract}

		\emph{Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.

@@

	In crystallography, cell geometry is conventionally encoded in a six-dimensional parameter space 
	\ensuremath{\HVI{}=(a,b,c, \alpha, \beta, \gamma).}\footnote{``H'' was chosen to
		honor early French crystallographer Ren\'e Just Ha\"uy.} But \HVI{} is not a metric space
	(where distances can be simply defined), and \HVI{} is not a vector space (where 
	objects can be added and subtracted). In the same sense, symmetry operations
	have no simple representations in \HVI{}.

@@

\begin{table}
	\label{tbl_spaces}
	\centering
	\caption{Crystallographic spaces with their scalar definitions and analytical purposes}
	\renewcommand{\arraystretch}{1.3}
	\begin{tabular}{|c|l|p{7cm}|}
		\hline
		\textbf{Space} & \textbf{Scalars / Coordinates} & \textbf{Purpose} \\

\hline
$\HVI$ &
\parbox[t]{5.5cm}{
	$a,\, b,\, c \in \mathbb{R}_{>0}$ — edge lengths\\
	$\alpha,\, \beta,\, \gamma \in (0^\circ, 180^\circ)$ — opposing angles\\
	Additionally, the sum of the angles must be less than 
	$360^\circ$ and must obey the triangle inequality
} &
Composite parameter space described as 
$\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$. 
The length dimensions are unconstrained, while the angles are 
geometrically restricted. Useful for conventional lattice 
descriptions, but limited in symmetry representation and 
metric-based classification. \\

\hline		
$\mathbf{F^3}$ &
$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
Base vector space, a frame space (or frame bundle) in \RIII. The coordinates are the 3-space 
base vectors of a unit cell. Symmetry 
operations are enumerated using this space. \\
\hline

		$\mathbf{P^3}$ &
		\parbox[t]{5.5cm}{
			$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
			$\Rightarrow$\,\\ 
			\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha$\\
		\hspace*{0.5cm}	$|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta$\\
		\hspace*{0.5cm}	$|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$ \\
		also known as: \\
		\hspace*{0.5cm}	\ensuremath{(p1,p2,p3)}. 
		} &
		Polar coordinate space defined by the map $\FIII \rightarrow \PIII$. Represents each base vector by magnitude and opposing angle, decomposed into directional components. Compact and smooth alternative to $\HVI$. \PIII{} provides a simple
		and intuitive space for representing and comparing
		 large numbers of unit cells. (This work) \\
		\hline

$\mathbf{S^6}$ &
\parbox[t]{5.5cm}{
	$s_1 = \vec{b} \cdot \vec{c},\, s_2 = \vec{a} \cdot \vec{c},\, s_3 = \vec{a} \cdot \vec{b}$\\
	$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\ \\
	\quad where $\vec{d} = -\vec{a} - \vec{b} - \vec{c}$
} &
A space used for Selling/Delone reduction. The scalars are derived from dot products between lattice vectors. Used for lattice classification described by Delaunay and for distance computation \cite{Andrews2019b}. 24 Bravais types were defined by citeasnoun{Delone1975} in \SVI{}, but 
described using pictures showing the special relationships. \citeasnoun{andrews2023measuring} revised the table of Delone to
include the \SVI vector definitions of the Bravais types.\\

		\hline
		$\mathbf{G^6}$ &
		\parbox[t]{5.5cm}{
			$g_1 = \vec{a} \cdot \vec{a},\, g_2 = \vec{b} \cdot \vec{b},\, g_3 = \vec{c} \cdot \vec{c}$\\
			$g_4 = 2\vec{b} \cdot \vec{c},\, g_5 =2 \vec{c} \cdot \vec{a},\, g_6 = 2\vec{a} \cdot \vec{b}$
		} &
		A space based on the metric tensor (and thus dot products 
		of the basis vectors of the unit cell) and used for Niggli reduction and lattice distance computation. \cite{Andrews2014}
		The International Tables for Crystallography describe the 44
		Bravais types that result from Niggli reduction using
		algebraic expression in \GVI{} (but stated in terms of scalars
		rather than manifolds in \GVI{}) \cite{ITC_VolumeA_2016}. \\
		\hline

\CIII &
\parbox[t]{5.5cm}{
	from the \SVI{} scalars, $s1,...,s6$\\
	$c1=(s1,s4), c2=(s2,s5), c3=(s3,s6)$\\
	where (?,?) denotes a complex number
} &
\CIII{} was created to associate related scalars after the
indication of \citeasnoun{Delone1975} concerning the ``opposite''
scalars in \SVI{}. Symmetry and Selling/Delaunay reduction have
simpler forms in \CIII{} than in \SVI{}. \cite{andrews2023complex}\\
\hline

		$\mathbf{DC7u}$ &
		\parbox[t]{5.5cm}{
			$d_1,\, d_2,\, d_3$ — cell edge lengths\\
			$d_4,\, d_5,\, d_6$ — shorter face diagonals' lengths\\
			$u$ — shortest body diagonal
		} &
		Unsorted Dirichlet Cell space derived from Niggli-reduced cells. Defined by the map from lattice geometry to a 7-dimensional vector of distances. Fully invertible and smooth; used for rapid lattice distance computation. \cite{bernstein2023invertible}\\
		\hline

	\end{tabular}
\end{table}

@@

\end{tikzpicture}
\label{fig_spaces}
\caption{Derived relationships among crystallographic spaces. 
	Base vector space \FIII{} gives rise to three scalar spaces:
	metric tensor space \GVI{}, Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}.
	Each transforms into a reduced or classification space: 
	$\mathbf{DC7u}$ from \GVI{}, \CIII{} from \SVI{}, and \PIII{} from $\mathbf{H^6}$.}
\end{figure}

@@

\citeasnoun{Delone1975} repeatedly pointed out the relationship of the
``opposite'' scalars in the tetrahedron representations of the Selling
scalars. Examining those relationships show that the opposite terms 
are pairs where a term involving a is opposite a term involving $\alpha$, etc.

Extending the observation of \citeasnoun{Delone1975}, the idea occurs
to make a triple of variables of those pairs. Since each pair has
one length and one angle, we considered each pair as the definition of
a polar coordinate space. The result is \PIII{}, a space of 3 2-dimensional polar coordinates. See Table \ref{tbl_spaces} for the
formal description.  Two uses immediately are obvious.

The first uses the 2-dimensional character of the coordinates of \PIII{}. They
allow simple to plot representations of collections of unit cells. 
See Figure \ref{PlotPolar_raw}, Figure \ref{PlotPolar_Niggli}, and Figure \ref{PlotPolar_Delone}.

\begin{figure}
\includegraphics[width=\columnwidth]{PlotPolar_raw}
\label{PlotPolar_raw}
	\caption{100 random primitive triclinic unit cells}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{PlotPolar_Niggli}
	\label{PlotPolar_Niggli}
	\caption{The same100 random primitive triclinic unit cells, Niggli-reduced}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{PlotPolar_Delone}
	\label{PlotPolar_Delone}
	\caption{The same 100 random primitive triclinic unit cells, Selling-reduced}
\end{figure}

%\[
%(|\vec{x}|\cos\theta,\, |\vec{x}|\sin\theta) \in \mathbb{R}^2
%\]

%which preserves the Euclidean magnitude and implicitly encodes angular orientation. While the transformation is nonlinear in angle, the resulting structure allows straightforward computation of relative orientation, cluster proximity, and shape similarity through standard Euclidean metrics.

The second use is allowing the calculation of Euclidean distances. Because \PIII{} can be represented as 3 xy planes, the Euclidean difference between
any two unit cells described as \PIII{} can be easily computed. Thus use
contrasts with \HVI{}, where incommensurate lengths and angles do not
lend themselves to a distance calculation.

@@

While derived from the same foundational cell parameters as $\mathbf{H^6}$, the space $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in conventional cell representations. Compared with $\mathbf{C^3}$, which seeks symmetry-reduced coordinates derived from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic. It provides a useful intermediary: more interpretable than metric tensors or other spaces, less abstract than \SVI{} or \GVI{}.

	\section{Notation}

	\section{Matrices of boundary transformations}

	\subsection{Derivation}

	\subsection{Example P++ implementation}

	\subsection{Example of use}

	\section{Graphical display of projections}

	\section{Scalar Reduction and Canonical Presentation in Projected Vector Triplets}

	\subsection{1. Standard Presentation of Projected Vectors}

	Let a set of three projected vectors be given:

	\[
	(p_1, p_2, p_3), \quad p_i \in \mathbb{R}^2
	\]

	Each \( p_i \) is constructed from a lattice vector via magnitude and opposing angle:

	\[
	p_i = |\vec{v}_i| (\cos \theta_i, \sin \theta_i)
	\]

	The \textbf{Standard Presentation} algorithm assigns a canonical ordering and orientation to the triplet:

	\begin{enumerate}
		\item \textbf{Magnitude Sort:} Reorder vectors so that

		\[
		|p_1| \leq |p_2| \leq |p_3|
		\]

		using Euclidean norms \( |p_i| = \sqrt{x_i^2 + y_i^2} \). Ties are preserved up to a numerical tolerance \( \delta \).

\item \textbf{Directional Coherence:} If the \( x \)-components of the triplet are not all of consistent sign, flip the necessary vectors by negating only the \( x_i \) component:

\[
p_i \to (-x_i,\ y_i)
\]

This maps quadrant I vectors into quadrant II, preserving their placement in the upper half-plane and maintaining angular conventions typical in crystallographic analysis.

\[
p_i \to (-x_i,\ y_i)
\]

to align the triplet into a directionally consistent state. If all \( x_i < 0 \), the configuration is already coherent and requires no change.

		\[
		p_i \to (-x_i, y_i)
		\]

		for any two vectors, preserving chirality.

		\item \textbf{Return Ordered Triplet:} Output the permuted and flipped triplet with consistent magnitude order and cosine directionality.
	\end{enumerate}

	\vspace{1em}

	\subsection{2. Scalar Projection Interaction and Cost Function}

	For any pair of projected vectors \( p_i, p_j \in \mathbb{R}^2 \), define the scalar projection interaction (or directional affinity):

	\[
	\mu_{ij} = p_i \cdot p_j = x_i x_j + y_i y_j
	\]

	The \textbf{scalar coupling cost} is then defined over the triplet as:

	\[
	C' = \frac{|\mu_{12}| + |\mu_{13}| + |\mu_{23}|}{p_1 \cdot p_1 + p_2 \cdot p_2 + p_3 \cdot p_3}
	\]

	This cost measures the normalized sum of pairwise directional alignment. Lower values of \( C' \) indicate that the projected vectors are less directionally coupled — i.e., more orthogonal or dispersed.

	\vspace{1em}

	\subsection{3. Scalar Reduction Algorithm}

	The goal of scalar reduction is to transform the triplet \( (p_1, p_2, p_3) \) into an algebraically simpler form — one with lower scalar coupling cost \( C' \) — while preserving lattice identity (e.g., volume and reconstructability).

	The reduction proceeds iteratively:

	\begin{enumerate}
		\item \textbf{Initialization:}
		Apply standard presentation. Compute initial cost \( C'_0 \) and physical volume \( V_0 > 0 \).

		\item \textbf{Pairwise Projection Subtractions:}
		For each unordered pair \( (p_i, p_j),\ i \neq j \), compute the scalar coefficient:

		\[
		\lambda = \frac{p_i \cdot p_j}{p_j \cdot p_j}
		\]

		Then define the candidate vector:

		\[
		p_i' = p_i - \lambda p_j
		\]

		\item \textbf{Evaluate Candidate:}
		Apply standard presentation to updated triplet. Recompute cost \( C'_{\text{new}} \) and volume \( V_{\text{new}} \).

		\item \textbf{Acceptance Criteria:}
		Accept the transformation if:

		\[
		C'_{\text{new}} + \delta < C' \quad \text{and} \quad V_{\text{new}} > 0
		\]

		\item \textbf{Repeat:}
		Continue until no further accepted transformations lower the cost.

		\item \textbf{Output:}
		Return the reduced triplet \( (p_1, p_2, p_3)_{\text{red}} \) with minimal scalar cost.
	\end{enumerate}

	\vspace{1em}

	\subsection{4. Interpretation of Cost and Reduction}

	The term "reduction" refers to minimizing the scalar coupling among vectors — reducing directional alignment that obscures presentation clarity. The cost \( C' \) is a rigorously defined metric for this alignment. Transformations reduce \( C' \) while preserving:
	\begin{itemize}
		\item The vector magnitudes and their reconstructable relationships
		\item The unit cell's volume and crystallographic identity
		\item The triplet’s algebraic validity in scalar projection space
	\end{itemize}

	Reduction leads to a cleaner, more canonical representation of the original structure — just as Niggli reduction provides a minimal cell in metric tensor space, scalar reduction provides a minimal configuration in projected vector space.

>>>>

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
%\usepackage{calligra}
\usepackage{longtable}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
%\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}
\usepackage[normalem]{ulem}
% Redefine \underline to use \textit
\renewcommand{\underline}[1]{\textit{#1}}

@@

% lca IUCr id IUCr6401
% HJB IUCr id IUCr6484
% NKS IUCr ID: IUCr7572
% lca ORCID  0000-0002-4451-1641
% HJB ORCID 0000-0002-0517-8532
% NKS ORCID 0000-0003-2786-6552

@@

% Crystallographic space macros
\newcommand{\HVI}{\ensuremath{\mathbf{H}^{6}}}
\newcommand{\GVI}{\ensuremath{\mathbf{G}^{6}}}
\newcommand{\SVI}{\ensuremath{\mathbf{S}^{6}}}
\newcommand{\PIII}{\ensuremath{\mathbf{P}^{3}}}
\newcommand{\FIII}{\ensuremath{\mathbf{F}^{3}}}
\newcommand{\BIV}{\ensuremath{\mathbf{B}^{4}}}
\newcommand{\CIII}{\ensuremath{\mathbf{C}^{3}}}
\newcommand{\RIII}{\ensuremath{\mathbb{R}^{3}}}

@@

	\title{\PIII, linearizing unit cell parameters}
	\shorttitle{properties of \PIII}

@@

\aff[a]{Ronin Institute for Independent Scholarship 2.0, USA}
\aff[b]{Ronin Institute for Independent Scholarship 2.0, USA}

@@

	\begin{abstract}
		The space \PIII{} is introduced, derived from unit cell axial lengths and interaxial angles. \PIII enables linearization of unit cell parameters using three polar coordinate bases.

	\PIII{} serves as a compact, 
	interpretable alternative to more abstract spaces such as \GVI{} and \SVI{}.

@@

	In crystallography, unit cell geometry is conventionally encoded in a six-dimensional parameter space 
	\ensuremath{\HVI{}=(a,b,c, \alpha, \beta, \gamma).}\footnote{``H'' was chosen to
		honor early French crystallographer Ren\'e Just Ha\"uy.} However, \HVI{} is not a metric space
	(where distances can be simply defined), nor is \HVI{} a vector space (where 
	objects can be added and subtracted). In the same sense, symmetry operations
	have no simple representations in \HVI{}. In consequence, \HVI{} has no
	simple measure of the ``difference'' between two lattices; how can one 
	compare a difference of 1 \AA{}ngstrom unit vs. one angular degree?

@@

We propose a ``linearized'' space, \PIII{}, derived from \HVI{}, but
having all measures in \AA{}ngstrom units (or other chosen linear measures.)

\renewcommand{\arraystretch}{1.3}

\begin{longtable}{@{}c@{\quad}l@{\quad}p{7cm}@{}}
	\caption{Crystallographic spaces with their scalar definitions and analytical purposes}
	\label{tbl_spaces} \\

	\toprule
	\textbf{Space} & \textbf{Scalars / Coordinates} & \textbf{Purpose} \\
	\midrule
	\endfirsthead

	\multicolumn{3}{@{}c@{}}{\textit{Table \thetable\ continued}} \\
	\toprule
	\textbf{Space} & \textbf{Scalars / Coordinates} & \textbf{Purpose} \\
	\midrule
	\endhead

	\midrule
	\multicolumn{3}{@{}c@{}}{\textit{Continued on next page}} \\
	\endfoot

	\bottomrule
	\endlastfoot

	$\HVI$ &
	\parbox[t]{5.5cm}{
		$a,\, b,\, c \in \mathbb{R}_{>0}$ — edge lengths\\
		$\alpha,\, \beta,\, \gamma \in (0^\circ, 180^\circ)$ — opposing angles\\

	} &
	Composite geometric parameter space $\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$. Lengths are unconstrained; angles are geometrically restricted. The sum of the angles must be less than $360^\circ$ and obey the triangle inequality \HVI{} is useful for conventional lattice descriptions, but limited in symmetry representation and metric-based classification. \\
	\midrule
	\FIII{} &
	$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
	Base vector space in \RIII. Coordinates are the three-space base vectors of a unit cell. Lattices are defined using the
	components of \FIII{}, and translational symmetry is described 
	in \FIII{}. \\	
	\midrule

	\BIV{} &
	\ensuremath{\vec{a},\, \vec{b},\, \vec{c},\, (-\vec{a}-\vec{b}-\vec{c}) \text{ in } \mathbb{R}^3} &
	Base vector space in \FIII{} plus the negative sum of the three base vectors. 
	This set of vectors is used in the definition of \SVI{}, which
	is the space of Selling/Delone reduction.\\
	\midrule

	\PIII{} &
	\parbox[t]{5.5cm}{
		$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
		$\Rightarrow$\\
		\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha)$\\
		\hspace*{0.5cm}$(|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta)$\\
		\hspace*{0.5cm}$(|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$\\
		also known as $(p_1, p_2, p_3)$
	} &
	Polar coordinate space defined in terms to unit cell axial
	lengths and their opposing angle. It is a compact and smooth alternative to $\HVI$. Useful for comparing unit cells. (This work) \\
		\midrule

	\SVI{} &
	\parbox[t]{5.5cm}{
		$s_1 = \vec{b} \cdot \vec{c},\, s_2 = \vec{a} \cdot \vec{c},\, s_3 = \vec{a} \cdot \vec{b}$\\
		$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\
		\quad where $\vec{d} = -\vec{a} - \vec{b} - \vec{c}$
	} &
	Used for Selling/Delone reduction. The scalar components are dot products between lattice vectors. \SVI{} supports Bravsi lattice classification. 24 Bravais types were defined using the scalar components of \SVI{} by Delone (1975), later revised using \SVI{} \cite{andrews2023measuring}. \\
		\midrule

	\GVI{} &
	\parbox[t]{5.5cm}{
		$g_1 = \vec{a} \cdot \vec{a},\, g_2 = \vec{b} \cdot \vec{b},\, g_3 = \vec{c} \cdot \vec{c}$\\
		$g_4 = 2\vec{b} \cdot \vec{c},\, g_5 = 2\vec{c} \cdot \vec{a},\, g_6 = 2\vec{a} \cdot \vec{b}$
	} &
	\GVI{} is based on the metric tensor and used for Niggli reduction and lattice distance computation \cite{Andrews2014}. The International Tables describe 44 Bravais types using scalar expressions derived from \GVI{} \cite{ITC_VolumeA_2016}. \\
		\midrule

\CIII{} &
	\parbox[t]{5.5cm}{
		From \SVI{} scalars: $s_1,\dots,s_6$\\
		$c_1 = (s_1, s_4),\ c_2 = (s_2, s_5),\ c_3 = (s_3, s_6)$\\
		where the parentheses denote a complex number
	} &
	\CIII{} associates scalar pairs identified by Delone as “opposites.” Symmetry and Selling/Delone reduction simplify in this space \cite{andrews2023complex}. \\
		\midrule

	$\mathbf{DC7u}$ &
	\parbox[t]{5.5cm}{
		$d_1,\, d_2,\, d_3$ — edge lengths\\
		$d_4,\, d_5,\, d_6$ — shortest face diagonals\\
		$u$ — shortest body diagonal
	} &
	$\mathbf{DC7u}$: unsorted Dirichlet Cell space derived from Niggli-reduced cells. Defined as a 7D vector of distances. Fully invertible and smooth; used for rapid lattice distance computation \cite{bernstein2023invertible}. \\

\end{longtable}

@@

\end{tikzpicture}
\label{fig_spaces}
\caption{Schematic relationships among crystallographic spaces. The base vector 
	space \FIII{} gives rise to three scalar spaces: metric tensor space \GVI{}, 
	Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}. 
	Each transforms into a reduced or classification space: $\mathbf{DC7u}$ from \GVI{}, 
	\CIII{} from \SVI{}, and \PIII{} from \HVI{}.}
\end{figure}

@@

\citeasnoun{Delone1975} repeatedly pointed out the relationship of the
``opposite'' scalars in the tetrahedron representations of the Selling
scalars. Examining those relationships shows that the opposite terms 
are pairs where a term involving $\mathbf{a}$ is opposite a term involving $\alpha$, {\it etc.}

Extending the observation of \citeasnoun{Delone1975}, the idea occurs
to make a triple of those pairs. Since each pair has
one length and one angle, we consider each pair as the definition of
a polar coordinate. The result is \PIII{}, a space of three two-dimensional polar coordinates. See Table \ref{tbl_spaces} for the
formal description.  Two uses immediately are obvious.

The first use applies the two-dimensional character of the coordinates of \PIII{}. They
allow simple-to-plot representations of collections of unit cells. Plotting
each of the coordinates of \PIII{} generates plots that are easily
interpreted as crystallographic axial length/opposite angle pairs. The
distance of a point has a length equal to the axial length, and the interaxisl
angle of the two other bases is illustrated on the plots. See Section  \ref{graphical}. Basically, together the three plots of the three coordinates are
simple projections of \HVI{}. See Section \ref{graphical}.

A second use is allowing the calculation of Euclidean distances between
pairs of unit cells. Because \PIII{} can be represented as three xy planes, the Euclidean difference between
any two unit cells described as \PIII{} can be easily computed. This use
contrasts with \HVI{}, where incommensurate lengths and angles do not
lend themselves to a distance calculation. See Section \ref{comparing}.

@@

While \PIII{} is derived from the same foundational cell parameters as $\mathbf{H^6}$, $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in the conventional cell representation. Compared with $\mathbf{C^3}$ \cite{andrews2023delone}, which derives coordinates from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic. It provides a useful intermediary: more interpretable than metric tensors or other spaces, but less abstract than \SVI{} or \GVI{}.

\section{Graphical display of projections}{
\label{graphical}
\PIII{}
allows simple-to-plot representations of collections of unit cells. 
See Figures \ref{PlotPolar_raw}, \ref{PlotPolar_Niggli}, and  \ref{PlotPolar_Delone} where 100 randomly chosen triclinic cells are plotted
(Figure \ref{PlotPolar_raw}) and the effects of unit cell reduction
are demonstrated (Figure \ref{PlotPolar_Niggli} and Figure \ref{PlotPolar_Delone} ).
}

\begin{figure}
    \begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_raw.png}
	\label{PlotPolar_raw}
	\caption{Scatter plots of 100 randomly selected
		 primitive triclinic unit cells represented in \PIII{}. Each of the 3 components of \PIII{}
		 is shown.}
	\end{center}
\end{figure}

\begin{figure}
    \begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_Niggli.png}
	\label{PlotPolar_Niggli}
	\caption{The same 100 random primitive triclinic unit cells shown in Figure \ref{PlotPolar_raw}, Niggli-reduced}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
	\includegraphics[width=\columnwidth]{PlotPolar_Delone.png}
	\label{PlotPolar_Delone}
	\caption{Scatter plots of 100 randomly selected
		primitive triclinic unit cells represented in \PIII{} as shown in Figure \ref{PlotPolar_raw}, Selling (Delaunay)-reduced}
	\end{center}
\end{figure}

\section{Comparing different unit cells}
\label{comparing}

In three-space, measuring the Euclidean distance between two points is just computing the square root
of the sum of the squares of the differences in the three coordinates. Similarly in \PIII{},
we compute distances from the sum of squared difference between the three
polar coordinates. But polar coordinates have a Euclidean measure from
their xy representation. So taking the six scalar values of two unit cells
presented in \PIII{}, we calculate a Euclidean distance by the ordinary
method.

%; CmdDists, calculate distances
%; Enter control variables and input vectors (type 'end' to finish):
%P 10 10 10  90 90 90
%p 10 10 10  90 90 91
%p 10 10 10.1 90 90 90
%end
%10.000    10.000    10.000    90.000    90.000    90.000
%14.142    30.113     8.479   147.466    56.504   144.893
%
%10.000    10.000    10.000    90.000    90.000    90.000
%14.142    29.916     8.564   147.586    56.504   144.621
%
%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
%; 1-2        0.175      19191.000      3.490         1-2         0.175      19191.000      3.490
%; 2-3        0.202      19191.000      4.028         1-3         0.100      19191.000      2.010
%
%
%;.......... input.cells ........................................ ordinals .....
%P   10.000    10.000    10.000    90.000    90.000    90.000     1
%P   10.000    10.000    10.000    90.000    90.000    91.000     2
%P   10.000    10.000    10.100    90.000    90.000    90.000     3
%;</pre>

%; CmdDists, calculate distances
%; Enter control variables and input vectors (type 'end' to finish):
%p 10 10 10   90 90 90
%p 10 10 10   90 90 91
%p 10 10 11   90 90 90
%end
%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
%; 1-2        3.490        0.203        3.490         1-2         0.175       26.959        3.490
%; 2-3        21.288       0.872       21.288         1-3         1.000       26.896       21.000
%
%
%;.......... input.cells ........................................ ordinals .....
%P   10.000    10.000    10.000    90.000    90.000    90.000     1
%P   10.000    10.000    10.000    90.000    90.000    91.000     2
%P   10.000    10.000    11.000    90.000    90.000    90.000     3
%;</pre>

Let us compare two cells derived by modifying a primitive unit cell to form two slightly modified cells, one by changing $\gamma$, and one by changing $c$.
\begin{enumerate}
	\item P 10 10 10  90 90 90
	\item P 10 10 10  90 90 91
	\item P 10 10 10.1  90 90 90
\end{enumerate}
The Euclidean distances between these cells are:
\begin{itemize}
	\item 1-2: 0.175
	\item 1-3: 0.100
	\item 2-3: 0.202
\end{itemize}

In this example, a 1.0° change in $\gamma$ yields a Euclidean distance comparable to a 0.1 Å change in the c-axis length.
%
%\section{Scalar Reduction and Canonical Presentation in Projected Vector Triplets}
%
%\subsection{1. Standard Presentation of Projected Vectors}
%	
%	Let a set of three projected vectors be given:
%	
%	
%	\[
%	(p_1, p_2, p_3), \quad p_i \in \mathbb{R}^2
%	\]
%	
%	
%	Each \( p_i \) is constructed from a lattice vector via magnitude and opposing angle:
%	
%	
%	\[
%	p_i = |\vec{v}_i| (\cos \theta_i, \sin \theta_i)
%	\]
%	
%	
%	
%	The \textbf{Standard Presentation} algorithm assigns a canonical ordering and orientation to the triplet:
%	
%	\begin{enumerate}
%		\item \textbf{Magnitude Sort:} Reorder vectors so that
%		
%		
%		\[
%		|p_1| \leq |p_2| \leq |p_3|
%		\]
%		
%		
%		using Euclidean norms \( |p_i| = \sqrt{x_i^2 + y_i^2} \). Ties are preserved up to a numerical tolerance \( \delta \).
%		
%\item \textbf{Directional Coherence:} If the \( x \)-components of the triplet are not all of consistent sign, flip the necessary vectors by negating only the \( x_i \) component:
%
%
%\[
%p_i \to (-x_i,\ y_i)
%\]
%
%
%This maps quadrant I vectors into quadrant II, preserving their placement in the upper half-plane and maintaining angular conventions typical in crystallographic analysis.
%
%
%
%\[
%p_i \to (-x_i,\ y_i)
%\]
%
%
%to align the triplet into a directionally consistent state. If all \( x_i < 0 \), the configuration is already coherent and requires no change.
%
%		
%		
%		\[
%		p_i \to (-x_i, y_i)
%		\]
%		
%		
%		for any two vectors, preserving chirality.
%		
%		\item \textbf{Return Ordered Triplet:} Output the permuted and flipped triplet with consistent magnitude order and cosine directionality.
%	\end{enumerate}
%	
%	\vspace{1em}
%	
%	\subsection{2. Scalar Projection Interaction and Cost Function}
%	
%	For any pair of projected vectors \( p_i, p_j \in \mathbb{R}^2 \), define the scalar projection interaction (or directional affinity):
%	
%	
%	\[
%	\mu_{ij} = p_i \cdot p_j = x_i x_j + y_i y_j
%	\]
%	
%	
%	
%	The \textbf{scalar coupling cost} is then defined over the triplet as:
%	
%	
%	\[
%	C' = \frac{|\mu_{12}| + |\mu_{13}| + |\mu_{23}|}{p_1 \cdot p_1 + p_2 \cdot p_2 + p_3 \cdot p_3}
%	\]
%	
%	
%	
%	This cost measures the normalized sum of pairwise directional alignment. Lower values of \( C' \) indicate that the projected vectors are less directionally coupled — i.e., more orthogonal or dispersed.
%	
%	\vspace{1em}
%	
%	\subsection{3. Scalar Reduction Algorithm}
%	
%	The goal of scalar reduction is to transform the triplet \( (p_1, p_2, p_3) \) into an algebraically simpler form — one with lower scalar coupling cost \( C' \) — while preserving lattice identity (e.g., volume and reconstructability).
%	
%	The reduction proceeds iteratively:
%	
%	\begin{enumerate}
%		\item \textbf{Initialization:}
%		Apply standard presentation. Compute initial cost \( C'_0 \) and physical volume \( V_0 > 0 \).
%		
%		\item \textbf{Pairwise Projection Subtractions:}
